---
title: "Final project presentation"
author: "Youcef Djoudi"
date: "11/29/2023"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Crime statistics data set for the City of Houston
- In this project I processed data set of 2022 crime statistics for the City of Houston. The data set is posted in the City of Houston Official website
- The NIBRS Data is updated monthly in one aggregated Excel spreadsheet, NIBRS stands for National Incident-Based Reporting System
- the data on the Excel spread sheet only shows the police calls that resulted in filing a report, keep in mind that most police calls doesn't warrant a police report for example Alarm calls, traffic stops, trespassers, disturbances.
- This data does not account for unreported crimes in the City of Houston.
- The data shows incident numbers and organized by Zip codes, Police Districts and Beats, street and block numbers.


## Data dictionary
- the data contains 249912 rows and 16 variables.
- Occurrence Date: date of the incident.
- Occurrence Hour: hour when the incident occurred, recorded in military time.
- NIBRS Description: offense type.
- Offense Count: number of offenses committed.
- Beat: beat where the incident occurred.
- Premise : location type where the incident occurred.
- Block Range: street block number where the incident occurred.
- Street Name: street name where the incident occurred.
- Street Type: street type (Dr, St, Ave)
- Suffix [Streets]: N, S, W, E.



```{r}
library(readxl)
library(tidyverse)
library(dplyr)

file_path <- "NIBRSPublicViewDec22.xlsx"

crime_stats <- read_excel(file_path)

```
## Glimpse of Dataset
```{r}
glimpse(crime_stats)
```

## Zip Code summary

- The minimum Zip Code 36830 is outside the range of Zip Codes in Houston.
- For the Zip Code plots I input the right Zip Codes for the city of Houston to avoid bad data. 
- I plotted incidents per Zip Codes on three different plots so it is easier to read.


```{r}
summary(crime_stats$ZIPCode)
```
## Number of incidents per Zip Codes (77002 to 77036)
```{r}
library(ggplot2)

# Filter the data to include only ZIP codes between 77001 and 77036
filtered_data_1 <- crime_stats %>%
  filter(ZIPCode >= 77002, ZIPCode <= 77036)

# Create a count plot with rotated x-axis labels
ggplot(filtered_data_1, aes(x = as.factor(ZIPCode))) +
  geom_bar(fill = "lightblue", color = "black") +
  labs(
    title = "Distribution of Crime Incidents by ZIP Code",
    x = "ZIP Code",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```



## Number of incidents per Zip Codes (77037 to 77071)

```{r}
# Filter the data to include only ZIP codes between 77037 and 77071
filtered_data_2 <- crime_stats %>%
  filter(ZIPCode >= 77037, ZIPCode <= 77071)

# Create a count plot with rotated x-axis labels
ggplot(filtered_data_2, aes(x = as.factor(ZIPCode))) +
  geom_bar(fill = "lightblue", color = "black") +
  labs(
    title = "Distribution of Crime Incidents by ZIP Code",
    x = "ZIP Code",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```



## Number of incidents per Zip Codes (77072 to 77099)

```{r}
# Filter the data to include only ZIP codes between 77071 and 77099
filtered_data_3 <- crime_stats %>%
  filter(ZIPCode >= 77072, ZIPCode <= 77099)

# Create a count plot with rotated x-axis labels
ggplot(filtered_data_3, aes(x = as.factor(ZIPCode))) +
  geom_bar(fill = "lightblue", color = "black") +
  labs(
    title = "Distribution of Crime Incidents by ZIP Code",
    x = "ZIP Code",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```


## Zip Codes with highest incidents
```{r}

zip_data <- crime_stats %>%
  group_by(ZIPCode) %>%
  summarise(Count = n())

zip_data_sorted <- zip_data %>%
  arrange(desc(Count))

# Select the top 10 ZIP codes
top_10_zip <- head(zip_data_sorted, 10)

# View the top 10 ZIP codes
print(top_10_zip)


```




## Number of reported incidents per month

```{r}
# Convert 'RMSOccurrenceDate' to a date object
crime_stats$RMSOccurrenceDate <- as.Date(crime_stats$RMSOccurrenceDate)

# Extract month from 'RMSOccurrenceDate'
crime_stats$Month <- format(crime_stats$RMSOccurrenceDate, "%Y-%m")

# Create a bar graph for the number of incidents per month
ggplot(crime_stats, aes(x = Month, fill=Month)) +
  geom_bar() +
  labs(title = "Number of Incidents per Month",
       x = "Month",
       y = "Number of Incidents") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```




## The most committed crimes in the city of Houston

```{r}
crime_counts <- crime_stats %>%
  group_by(NIBRSDescription) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

crime_counts %>% head(10)

```




## The least committed crimes in the City of Houston

```{r}
crime_counts <- crime_stats %>%
  group_by(NIBRSDescription) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

crime_counts %>% tail(10)
```




## Most committed crimes by Zip Codes

```{r}
crime_counts_by_zip <- crime_stats %>%
  group_by(NIBRSDescription, ZIPCode) %>%
  summarise(Count = n()) %>%
  arrange(ZIPCode, desc(Count)) %>%
  group_by(ZIPCode) %>%
  slice_head(n = 1)

crime_counts_by_zip

```



## Beats and districts 

- Police departments don't respond to calls by Zip codes. they respond to calls by districts
and beats.

- The city is divided to districts and each district is divided to beats.
- Beat is the area that's being patrolled
- Districts are divided between police stations throughout the city.
- Pdf file of districts and beats was obtained from the city of Houston website 
- The next slide display the city of Houston map per district and beat
- I patrolled 18 district from 2018 until 2021
- I investigate robbery cases in 1,2,9,10, and 17 districts 





```{r}
library(pdftools)

# Specify the path to your PDF file
pdf_path <- "hpd_beat_map.pdf"

# Extract images from the PDF file
pdf_images <- pdf_convert(pdf_path, format = "png", pages = 1)  # Change 'png' to the desired image format

# Display the extracted images
for (i in seq_along(pdf_images)) {
  png::readPNG(pdf_images[i]) |> grid::grid.raster()
}

if (!requireNamespace("png", quietly = TRUE)) {
  install.packages("png")
}
library(png)

# Read the PNG image
image <- readPNG("hpd_beat_map_1.png")

# Display the image
grid::grid.raster(image)

```





## Districts and beats map


```{r}
# Install and load the 'png' package if not already installed
if (!requireNamespace("png", quietly = TRUE)) {
  install.packages("png")
}
library(png)

# Read the PNG image
image <- readPNG("hpd_beat_map_1.png")

# Display the image
graphics::par(mar = rep(0, 4))  # Set margins to 0
graphics::plot(0:1, 0:1, type = 'n', xlab = '', ylab = '')  # Create an empty plot
graphics::rasterImage(image, 0, 0, 1, 1)  # Display the image


```



## What Beats have the highest incidents

```{r}
beat_crime_counts <- crime_stats %>%
  group_by(Beat) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

beat_crime_counts %>% head(10)

```



## Incidents per district

```{r}
# Extract district information from the 'Beat' variable
crime_stats <- crime_stats %>%
  mutate(District = substr(Beat, 1, 2))

# Group by district and count the number of incidents
incidents_by_district <- crime_stats %>%
  group_by(District) %>%
  summarise(IncidentCount = n()) %>%
  arrange(desc(IncidentCount))

# Create a bar graph for the number of incidents per district
ggplot(incidents_by_district, aes(x = District, y = IncidentCount, fill = District)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Incidents per District",
       x = "District",
       y = "Number of Incidents") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```




## Districts with highest incidents

```{r}
incidents_by_district %>% head(10)

```



## Districts with the lowest incident

```{r}
incidents_by_district %>% tail(10)

```





## Murder count of the year 2022

```{r}
murder_data <- crime_stats %>%
  filter(NIBRSDescription =="Murder, non-negligent")
murder_data

```




## Zip Codes with highest murder rate

```{r}
highest_murder_count_zip <- murder_data %>%
  group_by(ZIPCode) %>%
  summarise(MurderCount = n()) %>%
  arrange(desc(MurderCount))

highest_murder_count_zip %>% head(10)

```





## Zip Codes with lowest murder rate

```{r}
highest_murder_count_zip %>% tail(10)

```




## Plot murder rate by Zip Codes (77002 to 77036)

```{r}
filtered_data <- subset(murder_data, ZIPCode >= 77002 & ZIPCode <= 77036)

# Create a bar graph for murder counts by ZIPCode
ggplot(filtered_data, aes(x = as.factor(ZIPCode))) +
  geom_bar() +
  labs(title = "Murder Counts by ZIP Code",
       x = "ZIP Code",
       y = "Murder Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```



## Murder rate plot by Zip Codes (77037 to 77071)

```{r}
filtered_data <- subset(murder_data, ZIPCode >= 77037 & ZIPCode <= 77071)

# Create a bar graph for murder counts by ZIPCode
ggplot(filtered_data, aes(x = as.factor(ZIPCode))) +
  geom_bar() +
  labs(title = "Murder Counts by ZIP Code",
       x = "ZIP Code",
       y = "Murder Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```



## Murder rate plot by Zip Codes (77072 to 77099)

```{r}
filtered_data <- subset(murder_data, ZIPCode >= 77072 & ZIPCode <= 77099)

# Create a bar graph for murder counts by ZIPCode
ggplot(filtered_data, aes(x = as.factor(ZIPCode))) +
  geom_bar() +
  labs(title = "Murder Counts by ZIP Code",
       x = "ZIP Code",
       y = "Murder Count") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```





## Murder distribution per month
```{r}
# Convert 'RMSOccurrenceDate' to a date object
murder_data$RMSOccurrenceDate <- as.Date(murder_data$RMSOccurrenceDate)

# Extract month from 'RMSOccurrenceDate'
murder_data$Month <- format(murder_data$RMSOccurrenceDate, "%Y-%m")

# Create a bar graph for murder counts per month
ggplot(murder_data, aes(x = Month, fill=Month)) +
  geom_bar() +
  labs(title = "Distribution of Murders per Month",
       x = "Month",
       y = "Murder Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```




## Murder ditribution per district 
```{r}
# Create a subset for murder incidents
murder_data <- crime_stats %>% filter(NIBRSDescription == "Murder, non-negligent")

# Extract district information from the 'Beat' variable
murder_data <- murder_data %>%
  mutate(District = substr(Beat, 1, 2))

# Group by district and count the number of incidents
murder_by_district <- murder_data %>%
  group_by(District) %>%
  summarise(IncidentCount = n()) %>%
  arrange(desc(IncidentCount))

# Print the resulting data
print(murder_by_district)
```




## Murder rate per district plot

```{r}
ggplot(murder_by_district, aes(x = District, y = IncidentCount, fill = District)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Murders Incidents per District",
       x = "District",
       y = "Number of Murder Incidents") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```



## Number of robberies committed in 2022

```{r}

robbery_data<-crime_stats %>% filter(NIBRSDescription=="Robbery")
robbery_data

```





## Zip Codes with the highest robberies 

```{r}
robbery_by_zip <- robbery_data %>%
  group_by(ZIPCode) %>%
  summarise(IncidentCount = n()) %>%
  arrange(desc(IncidentCount))
robbery_by_zip %>% head(10)

```





## Zip Codes with the lowest robberies

```{r}
robbery_by_zip %>% tail(10)
```



## Robberies by beat

```{r}
robbery_by_beat <- robbery_data %>%
  group_by(Beat) %>%
  summarise(IncidentCount = n()) %>%
  arrange(desc(IncidentCount))
robbery_by_beat %>% head(10)

```



## Beats with lowest robberies

```{r}
robbery_by_beat %>% tail(10)

```




## Robberies distribution by month

```{r}
# Convert 'RMSOccurrenceDate' to a date object
robbery_data$RMSOccurrenceDate <- as.Date(robbery_data$RMSOccurrenceDate)

# Extract month from 'RMSOccurrenceDate'
robbery_data$Month <- format(robbery_data$RMSOccurrenceDate, "%Y-%m")

# Create a bar graph for the number of robberies per month
ggplot(robbery_data, aes(x = Month, fill=Month)) +
  geom_bar() +
  labs(title = "Number of Robberies per Month",
       x = "Month",
       y = "Number of Robberies") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



## Districts with the highest robberies 

```{r}
# Extract district information from the 'Beat' variable
robbery_data <- robbery_data %>%
  mutate(District = substr(Beat, 1, 2))

# Group by district and count the number of incidents
robbery_by_district <- robbery_data %>%
  group_by(District) %>%
  summarise(IncidentCount = n()) %>%
  arrange(desc(IncidentCount))

# Print the resulting data
print(robbery_by_district)

```




## Districts with lowest robberies

```{r}
robbery_by_district %>% tail(10)

```



## Robberies per district plot

```{r}

ggplot(robbery_by_district, aes(x = District, y = IncidentCount, fill = District)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Robberies per District",
       x = "District",
       y = "Number of Robberies") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



## Incidents by premise 

```{r}
# Group by premise and count the number of incidents
incidents_by_premise <- crime_stats %>%
  group_by(Premise) %>%
  summarise(IncidentCount = n()) %>%
  arrange(desc(IncidentCount))

# Print the resulting data
print(incidents_by_premise)

```




## Incidents distribution per hour of the day

```{r}
# Group by the hour of the day and count the number of incidents
incidents_by_hour <- crime_stats %>%
  group_by(RMSOccurrenceHour) %>%
  summarise(IncidentCount = n()) %>%
  arrange(RMSOccurrenceHour)

# Print the resulting data
#print(incidents_by_hour)

# Create a bar graph for the number of incidents by the hour of the day
ggplot(incidents_by_hour, aes(x = RMSOccurrenceHour, y = IncidentCount, fill = factor(RMSOccurrenceHour))) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Incidents by Hour of the Day",
       x = "Hour of the Day",
       y = "Number of Incidents") +
  theme(axis.text.x = element_text(angle = 0))

```




## Heat map per district

```{r}
# Extract district information from the 'Beat' variable
crime_stats <- crime_stats %>%
  mutate(District = substr(Beat, 1, 2))

# Group by district and hour, count the number of incidents
incidents_by_district_hour <- crime_stats %>%
  group_by(District, RMSOccurrenceHour) %>%
  summarise(IncidentCount = n())

# Create a heatmap
ggplot(incidents_by_district_hour, aes(x = RMSOccurrenceHour, y = District, fill = IncidentCount)) +
  geom_tile() +
  labs(title = "Crime Incidents Heatmap by District",
       x = "Hour of the Day",
       y = "District",
       fill = "Number of Incidents") +
  scale_fill_gradient(low = "green", high = "red") +  # You can adjust colors as needed
  theme_minimal()

```



## Conclusion
- from data analysis conducted on this data set I observed the following:
- 77036 Zip Codes has the highest rate of reported crimes
- 18 district has the highest incidents in the City of Houston
- Hour of the day that have highest incident is between 1200-1300
- Number of Homicides is 411 with 14 district having the highest murders rate (41), and
Zip code 77036 having the highest murder rate (21) per zip code
- Murder incidents have the highest distribution per month in June
- Robberies incident have the highest distribution per month in January
- Residential properties account for most of the reported incidents 
